Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-13 treestats_paper.tex
Minor edits

Naturally, the genealogical perspective on population genetics statistics has a long history.
The sequence of trees describing genealogical relationships as one moves along the genome
are also described by the Ancestral Recombination
Graph~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome}.
These genealogical trees can be fairly precisely inferred in nonrecombining regions
\citep{felsenstein2004inferring},
as for instance the mitochondria,
and substantial theoretical work has been done in the context of phylogenetics \citep{semple2003phylogenetics}.
% describing how each mutation gives a small piece of information about the tree at that site
% However, population genetics has the task of summarizing relationships across \emph{many} trees,
% each only partially known because mutations are sparse in the data.
The basic relationship between tree shape and summaries of genetic variation
is that if mutations are neutral then the expected number that occur on a given branch
is proportional to the length of that branch,
thus establishing a duality between genealogical and genetic summaries,
a fact used extensively in deriving properties of statistics in models of random mating
\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical}.
The fact that the duality applies regardless of the underlying demographic model was also used
\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,mcvean2002genealogical,ralph2019empirical},
but only recently have datasets made it possible to move beyond a few numerical summaries.

@@

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
In this paper, we have made an exploratory foray in this direction,
by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
between several subsets of the 1000 Genomes dataset,
in several inferred tree sequences.
It is not yet clear whether these time profiles tell us more about history
or about biases of the inference methods,
but either result would be useful.

@@

The tree sequence describes trees that change as one moves along the genome,
and so is a special case of a \emph{dynamic graph}
\citep[also called a \emph{graph timeline}][]{lacki2013reachability}.
Much of the work on dynamic graphs is focused on connectivity,
e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
but development of parallel algorithms for more general operations on large dynamic graphs
is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.

@@

\section*{Supplementary figures}

\begin{figure}
    \includegraphics{{figures/swept.10000.765.1e-09.diversity}.pdf}
    \caption{
        As in Figure~\ref{fig:sweep_duality}, but in a population of size 10,000.
        \label{sfig:sweep_duality_10K}
    }
\end{figure}

>>>>

Naturally, the genealogical perspective on population genetics statistics has a long history.
The sequence of trees describing genealogical relationships as one moves along the genome
are also described by the Ancestral Recombination
Graph~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome}.
These genealogical trees can be accurately inferred in nonrecombining regions
\citep{felsenstein2004inferring},
as for instance the mitochondria,
and substantial theoretical work has been done in the context of phylogenetics \citep{semple2003phylogenetics}.
% describing how each mutation gives a small piece of information about the tree at that site
% However, population genetics has the task of summarizing relationships across \emph{many} trees,
% each only partially known because mutations are sparse in the data.
The basic relationship between tree shape and summaries of genetic variation
is that if mutations are neutral then the expected number that occur on a given branch
is proportional to the length of that branch,
thus establishing a duality between genealogical and genetic summaries,
a fact used extensively in deriving properties of statistics in models of random mating
\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical}.
The fact that the duality applies regardless of the underlying demographic model was also used
\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,mcvean2002genealogical,ralph2019empirical},
but only recently have datasets made it possible to move beyond a few numerical summaries.

@@

% visualization and time decomposition adding a new dimension
\krt{this para needs to go, right, now that the related results are offloaded
into a new tex file? Perhaps say that we can do this sort of thing?}
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
In this paper, we have made an exploratory foray in this direction,
by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
between several subsets of the 1000 Genomes dataset,
in several inferred tree sequences.
It is not yet clear whether these time profiles tell us more about history
or about biases of the inference methods,
but either result would be useful.

@@

The tree sequence describes trees that change as one moves along the genome,
and so is a special case of a \emph{dynamic graph}, 
also called a \emph{graph timeline} \citep{lacki2013reachability}.
Much of the work on dynamic graphs is focused on connectivity,
e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
but development of parallel algorithms for more general operations on large dynamic graphs
is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.

@@

\clearpage
\newpage
\section*{Supplementary figures}

\begin{figure}[!h]
    \includegraphics{{figures/swept.10000.765.1e-09.diversity}.pdf}
    \caption{
        As in Figure~\ref{fig:sweep_duality}, but in a population of size 10,000.
        \label{sfig:sweep_duality_10K}
    }
\end{figure}
